---
title: "COVID-19: Your State is Not That Different"
author: Yiming (Paul) Li
date: '2020-03-29'
slug: covid-19-your-state-is-not-that-different
categories: []
tags: []
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = FALSE,
    warning = FALSE,
    message = FALSE,
    error = FALSE,
    collapse = TRUE,
    fig.align = "center",
    cache.comments = FALSE
)
```

```{r libraries}
library(tidyverse)
library(here)
library(lubridate)
library(janitor)
library(ggrepel)

theme_set(theme_light())
```

This post is inspired by the following sources:

* Kieran Healy's [replication](https://kieranhealy.org/blog/archives/2020/03/27/a-covid-small-multiple/) of John Burn-Murdoch's [small-multiple plot](https://www.ft.com/coronavirus-latest) of confirmed cases by country. The technique is useful to contrast a country's trajectory with other countries trajectories, putting things into perspective.
* The [log-log plot](https://aatishb.com/covidtrends/) of cumulative cases against new cases from last week, by [Aatish Bhatia](https://aatishb.com/) and [Minute Physics](https://www.youtube.com/watch?v=54XLXg4fYsc). The log-log scale and the stock vs (smoothed 7d) flow help us to see the exponential path each country was on, and make it easier to spot when the growth has passed the exponential growth phase and started to slow down.

I combine these two approaches to show that all 50 States and Washington DC in the United States are on the exponential growth path, and have not shown obvious signs of slowing down.


```{r, layout="l-body-outset", cache=TRUE}
us_states_raw <- read_csv("https://covidtracking.com/api/states/daily.csv") %>%
        janitor::clean_names()
us_states <- us_states_raw %>%
    mutate(date = lubridate::ymd(date)) %>%
    filter(date <= lubridate::today() - 1) %>%
    pivot_longer(positive:death, names_to = "measure", values_to = "count")
```

```{r}
state_name <- read_csv('
    "sname","state"
    "Alabama","AL"
    "Alaska","AK"
    "Arizona","AZ"
    "Arkansas","AR"
    "California","CA"
    "Colorado","CO"
    "Connecticut","CT"
    "Delaware","DE"
    "District of Columbia","DC"
    "Florida","FL"
    "Georgia","GA"
    "Hawaii","HI"
    "Idaho","ID"
    "Illinois","IL"
    "Indiana","IN"
    "Iowa","IA"
    "Kansas","KS"
    "Kentucky","KY"
    "Louisiana","LA"
    "Maine","ME"
    "Montana","MT"
    "Nebraska","NE"
    "Nevada","NV"
    "New Hampshire","NH"
    "New Jersey","NJ"
    "New Mexico","NM"
    "New York","NY"
    "North Carolina","NC"
    "North Dakota","ND"
    "Ohio","OH"
    "Oklahoma","OK"
    "Oregon","OR"
    "Maryland","MD"
    "Massachusetts","MA"
    "Michigan","MI"
    "Minnesota","MN"
    "Mississippi","MS"
    "Missouri","MO"
    "Pennsylvania","PA"
    "Rhode Island","RI"
    "South Carolina","SC"
    "South Dakota","SD"
    "Tennessee","TN"
    "Texas","TX"
    "Utah","UT"
    "Vermont","VT"
    "Virginia","VA"
    "Washington","WA"
    "West Virginia","WV"
    "Wisconsin","WI"
    "Wyoming","WY"
')

states_gt100 <-
    us_states %>%
    filter(measure == "positive") %>%
    group_by(state) %>%
    filter(date == max(date)) %>%
    filter(count >= 100) %>%
    pull(state)

states_cumulative <- 
    us_states %>%
    inner_join(state_name, by = "state") %>%
    filter(measure == "positive") %>%
    select(state, sname, date, count) %>%
    group_by(state) %>%
    mutate(
        increase_7d = count - lag(count, 7, order_by = date),
        increase_4d = count - lag(count, 4, order_by = date),
        increase_1d = count - lag(count, 1, order_by = date),
    ) %>%
    ungroup()

all_states_background <-
    states_cumulative %>%
    select(st = state, date, count, starts_with("increase_"))
    
endpoints <-
    states_cumulative %>%
    group_by(state) %>%
    filter(date == max(date)) %>%
    ungroup()

state_name_label <- 
    states_cumulative %>%
    group_by(state) %>%
    filter(date == max(date)) %>%
    ungroup() %>%
    mutate(
        count = 1, 
        increase_7d = max(increase_7d, na.rm = T) - 1e4,
        increase_4d = max(increase_4d, na.rm = T) - 1e4,
        increase_1d = max(increase_1d, na.rm = T) - 1e4,
    ) %>%
    select(state, sname, count, increase_7d)
```


```{r, fig.width=12, fig.height=20, dpi=300}
states_cumulative %>%
    ggplot(mapping = aes(x = count, y = increase_7d)) + 
    # The line traces for every country, in every panel
    geom_line(data = all_states_background, 
              aes(group = st),
              size = 0.15, color = "gray80") + 
    # The line trace in red, for the country in any given panel
    geom_line(color = "firebrick",
              lineend = "round") + 
    # The point at the end. Bonus trick: some points can have fills!
    geom_point(data = endpoints, 
               size = 1.1, 
               shape = 21, 
               color = "firebrick",
               fill = "firebrick2"
    ) + 
    # The country label inside the panel, in lieu of the strip label
    geom_text(data = state_name_label,
              mapping = aes(label = sname),
              vjust = "inward",
              hjust = "inward",
              fontface = "bold",
              color = "firebrick",
              size = 2.1) +
    # Log transform and friendly labels
    scale_x_log10(labels = scales::label_number_si()) + 
    scale_y_log10(labels = scales::label_number_si()) + 
    # Facet by country, order from high to low
    facet_wrap(~ reorder(state, -count), ncol = 5) + 
    labs(x = "Number of Confirmed Case (log10 scale)",
         y = "Number of Mew Confirmed Cases in the Last 7 Days (log10 scale)",
         title = "Exponential Growth of COVID-19 Confirmed Cases by US State",
         subtitle = paste("Data as of", format(max(us_states$date), "%A, %B %e, %Y")),
         caption = "Paul Yiming Li @paulymli / Data: https://covidtracking.com") +
    theme(plot.title = element_text(size = rel(1), face = "bold"),
          plot.subtitle = element_text(size = rel(0.7)),
          plot.caption = element_text(size = rel(0.7)),
          # turn off the strip label and tighten the panel spacing
          strip.text = element_blank(),
          panel.spacing.x = unit(-0.05, "lines"),
          panel.spacing.y = unit(0.3, "lines"),
          axis.text.y = element_text(size = rel(0.5)),
          axis.title.x = element_text(size = rel(1)),
          axis.title.y = element_text(size = rel(1)),
          axis.text.x = element_text(size = rel(0.5)),
          legend.text = element_text(size = rel(1)))
```

A few points to help with the interpretation of the graph, and a few caveats.

* Each day corresponds to a point on each of the graphs. Imaging a dot climbing up to the top right corner of each graph, with the large red dot representing the latest day.
* The gray background lines of each small-multiple panel are the growth paths of all other states.
* The 7-day period in the y axis is to smooth the daily fluctuations in new cases.
* If the growth of confirmed cases are exponential, the slope of log(confirmed cases) vs log(total new cases in the last 7 days) is 45 degrees. Therefore the fact that every state is on the 45-degree line suggests every state is on an exponential growth path.
* But 45-degree does not suggest all states share the same rate of growth, just that all states are on some exponential path.
* This graph only considers confirmed/detected cases, not the actually infected cases, which are only larger than detected cases. Differential ramp-up speed of testing across states is not taken into account.
